---
title: "R Notebook"
output: html_notebook
---

# load library
```{r}
library(tidyverse)
library(ggpubr)
library(phyloseq)
library(readxl)
```

# import data
```{r}
tax_file <- "data/metaphlan3.merged.abundance.profile.species.272S.tsv"
tax_s <- read_tsv(tax_file)
tax_s_l <-
    tax_s %>%
    rowwise() %>%
    mutate(Species=str_split(clade_name, "\\|s__")[[1]][2]) %>%
    ungroup() %>%
    pivot_longer(
        names_to="sample_id",
        values_to="Abundance",
        -c(clade_name, Species))

tax_file_p <- "data/metaphlan3.merged.abundance.profile.phylum.272S.tsv"
tax_p <- read_tsv(tax_file_p)
tax_p_l <-
    tax_p %>%
    rowwise() %>%
    mutate(Phylum=str_split(clade_name, "\\|p__")[[1]][2]) %>%
    ungroup() %>%
    pivot_longer(
        names_to="sample_id",
        values_to="Abundance",
        -c(clade_name, Phylum))
```

# read metadata
```{r}
samples_metadata <-
    read_tsv("metadata/samples_metadata_imm_272S_long.tsv")
    

paired_ID <- count(samples_metadata, ID) %>% filter(n==2) %>% select(ID) %>% as.list()

samples_metadata_paired <-
    samples_metadata %>% filter(ID %in% paired_ID$ID)

samples_metadata_ae <-
    read_tsv("metadata/samples_metadata_imm_ae_138S.tsv")
```

# merge tax profile and metadata
```{r}
tax_s_l_ae <- tax_s_l %>% inner_join(samples_metadata_ae)
tax_s_l_272S <- tax_s_l %>% inner_join(samples_metadata)
tax_s_l_268S <- tax_s_l %>% inner_join(samples_metadata_paired)
tax_p_l_268S <- tax_p_l %>% inner_join(samples_metadata_paired)
```

```{r}
dim(tax_s_l_272S)
```

## Species with immune binary group among D0 OWOB population
```{r}
samples_metadata_owob_group <-
    samples_metadata_ae %>%
    filter(TimePoint=="D0") %>%
    select(sample_id, ID, Vac_Group, OWOB, BMI,
           AUC_100_12800, M1_sVNT_10, M1_sVNT_200, M1_sVNT,
           RBD_binary_group, sVNT_binary_group, sVNT_60percent_binary_group) %>%
    distinct() %>%
    filter(!is.na(BMI)) %>%
    group_by(OWOB) %>%
    mutate(RBD_median=median(AUC_100_12800)) %>%
    mutate(RBD_median_group=case_when(
        AUC_100_12800 >= RBD_median ~ 1,
        AUC_100_12800 < RBD_median ~ 0))

tax_s_l_owob <- tax_s_l %>% right_join(samples_metadata_owob_group)
```

## Extended Figure 7

### OWOB1 and sVNT 60percent SinoVac
```{r}
tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group <-
    tax_s_l_owob %>%
    filter(OWOB==1) %>%
    filter(Vac_Group=="SinoVac") %>%
    select(sample_id, Species, Abundance, sVNT_60percent_binary_group) %>%
    group_by(Species) %>%
    mutate(Species_abun_sum=sum(Abundance)) %>%
    filter(!Species_abun_sum<=0.0) %>%
    select(-Species_abun_sum) %>%
    ungroup()

tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group_2 <-
    tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group %>%
    select(sample_id, sVNT_60percent_binary_group) %>%
    distinct() %>%
    mutate(Species="sVNT_60percent_binary_group") %>%
    rename(Abundance=sVNT_60percent_binary_group) %>%
    mutate(Abundance=as.numeric(Abundance))

tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group_w <-
    tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group_2 %>%
    bind_rows(tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group %>%
    select(sample_id, Species, Abundance)) %>%
    pivot_wider(names_from=sample_id, values_from=Abundance)
    
print(dim(tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group_w))
write_tsv(tax_s_l_d0_owob_1_sinovac_sVNT_60percent_binary_group_w,
          "tables/lefse/tax_s_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group_w.tsv")
```

### run LEfSe
```{bash}
lefse_format_input.py \
tables/lefse/tax_s_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group_w.tsv \
tables/lefse/tax_s_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group_w.lefse.in \
-c 2 -u 1 -o 1000000

lefse_run.py \
tables/lefse/tax_s_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group_w.lefse.in \
tables/lefse/tax_s_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group_w.lefse.out.tsv
```

### read LEfSe results
```{r}
lefse <-
  read_tsv(
    "tables/lefse/tax_s_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group_w.lefse.out.tsv",
    col_names=c("Species", "v1", "group", "LDA_score", "pvalue")) %>%
filter(!is.na(group)) %>%
mutate(LDA_score=case_when(
    group==0 ~ (-LDA_score),
    TRUE ~ LDA_score)) %>%
mutate(group=factor(group, levels=c("0", "1")))
```

### Extended figure 7
```{r}
p <- ggbarplot(
    lefse, x = "Species", y = "LDA_score",
    fill = "group",
    color = "white",
    sort.val = "asc",
    sort.by.groups = TRUE,
    ylab= FALSE,
    rotate = TRUE,
    legend.title = "sVNT 60percent binary group") +
ggtitle("LEfSe + D0 + OWOB1 + SinoVac + sVNT 60percent") +
ylab("LDA SCORE (log 10)") +
scale_fill_manual(breaks=c("0", "1"), labels=c("Low", "High"),
                  values=c("0"="#E69F00", "1"="#0072B2"))

print(p)
ggsave(
  "figures/Extended_figure_7-species_lefse_D0_OWOB1_sinovac_sVNT_60percent_binary_group.pdf",
  p, width=8, height=8)
```

## Extended figure 9d and 9h
### LEfSe analysis for extended figure 9d
```{r}
tax_s_lefse_sinovac_cluster <-
    tax_s_l_272S %>%
    filter(TimePoint=="D0" & Vac_Group=="SinoVac") %>%
    select(sample_id, Species, Abundance, SNV_jsd_cluster2) %>%
    distinct() %>%
    group_by(Species) %>%
    mutate(Species_abun_sum=sum(Abundance)) %>%
    filter(!Species_abun_sum<=0.0) %>%
    select(-Species_abun_sum) %>%
    ungroup()
    
tax_s_lefse_sinovac_cluster_2 <-
    tax_s_lefse_sinovac_cluster %>%
    select(sample_id, SNV_jsd_cluster2) %>%
    distinct() %>%
    mutate(Species="JSD_Cluster") %>%
    rename(Abundance=SNV_jsd_cluster2)
    
tax_s_lefse_sinovac_cluster_w <-
    tax_s_lefse_sinovac_cluster_2 %>%
    bind_rows(tax_s_lefse_sinovac_cluster %>%
    select(sample_id, Species, Abundance)) %>%
    pivot_wider(names_from=sample_id, values_from=Abundance)

print(dim(tax_s_lefse_sinovac_cluster_w))
write_tsv(tax_s_lefse_sinovac_cluster_w,
          "tables/lefse/tax_s_lefse_D0_sinovac_JSD_cluster.tsv")
```

### run LEfSe
```{bash}
lefse_format_input.py \
tables/lefse/tax_s_lefse_D0_sinovac_JSD_cluster.tsv \
tables/lefse/tax_s_lefse_D0_sinovac_JSD_cluster.lefse.in \
-c 2 -u 1 -o 1000000

lefse_run.py \
tables/lefse/tax_s_lefse_D0_sinovac_JSD_cluster.lefse.in \
tables/lefse/tax_s_lefse_D0_sinovac_JSD_cluster.lefse.out.tsv
```

```{r}
lefse <- read_tsv("tables/lefse/tax_s_lefse_D0_sinovac_JSD_cluster.lefse.out.tsv",
                  col_names=c("Species", "v1", "group", "LDA_score", "pvalue")) %>%
filter(!is.na(group)) %>%
mutate(LDA_score=case_when(
    group==2 ~ (-LDA_score),
    TRUE ~ LDA_score)) %>%
mutate(group=factor(group, levels=c("1", "2")))
```

### Extended figure 9d
```{r, fig.width=8, fig.height=10}
p <- ggbarplot(
    lefse, x = "Species", y = "LDA_score",
    fill = "group",
    color = "white",
    sort.val = "desc",
    sort.by.groups = TRUE,
    ylab= FALSE,
    rotate = TRUE,
    legend.title = "JSD Cluster") +
ggtitle("SinoVac JSD cluster") +
ylab("LDA SCORE (log 10)") +
scale_fill_manual(breaks=c("2", "1"), labels=c("Cluster 1", "Cluster 2"),
                  values=c("1"="#E69F00", "2"="#0072B2"))

print(p)
ggsave(
  "figures/Extended_figure_9d-species_lefse_D0_sinovac_JSD_cluster.pdf",
  p, width=8, height=8)
```


### LEfSe analysis for extended figure 9h
```{r}
tax_s_lefse_biontech_cluster <-
    tax_s_l_272S %>%
    filter(TimePoint=="D0" & Vac_Group=="BioNTech") %>%
    select(sample_id, Species, Abundance, BNT_jsd_cluster2) %>%
    distinct() %>%
    group_by(Species) %>%
    mutate(Species_abun_sum=sum(Abundance)) %>%
    filter(!Species_abun_sum<=0.0) %>%
    select(-Species_abun_sum) %>%
    ungroup()
    
tax_s_lefse_biontech_cluster_2 <-
    tax_s_lefse_biontech_cluster %>%
    select(sample_id, BNT_jsd_cluster2) %>%
    distinct() %>%
    mutate(Species="JSD_Cluster") %>%
    rename(Abundance=BNT_jsd_cluster2)
    
tax_s_lefse_biontech_cluster_w <-
    tax_s_lefse_biontech_cluster_2 %>%
    bind_rows(tax_s_lefse_biontech_cluster %>%
    select(sample_id, Species, Abundance)) %>%
    pivot_wider(names_from=sample_id, values_from=Abundance)

print(dim(tax_s_lefse_biontech_cluster_w))
write_tsv(tax_s_lefse_biontech_cluster_w,
          "tables/lefse/tax_s_lefse_D0_biontech_JSD_cluster.tsv")
```

### run LEfSe
```{bash}
/home/zhujie/.conda/envs/bioenv/bin/lefse_format_input.py \
tables/lefse/tax_s_lefse_D0_biontech_JSD_cluster.tsv \
tables/lefse/tax_s_lefse_D0_biontech_JSD_cluster.lefse.in \
-c 2 -u 1 -o 1000000

/home/zhujie/.conda/envs/bioenv/bin/lefse_format_input.py \
tables/lefse/tax_s_lefse_D0_biontech_JSD_cluster.lefse.in \
tables/lefse/tax_s_lefse_D0_biontech_JSD_cluster.lefse.out.tsv
```

```{r}
lefse <- read_tsv("tables/lefse/tax_s_lefse_D0_biontech_JSD_cluster.lefse.out.tsv",
                  col_names=c("Species", "v1", "group", "LDA_score", "pvalue")) %>%
filter(!is.na(group)) %>%
mutate(LDA_score=case_when(
    group==1 ~ (-LDA_score),
    TRUE ~ LDA_score)) %>%
mutate(group=factor(group, levels=c("2", "1")))
```


### Extended Figure 9h
```{r, fig.width=8, fig.height=12}
p <- ggbarplot(
    lefse, x = "Species", y = "LDA_score",
    fill = "group",
    color = "white",
    sort.val = "desc",
    sort.by.groups = TRUE,
    ylab= FALSE,
    rotate = TRUE,
    legend.title = "JSD Cluster") +
ggtitle("BioNTech JSD cluster") +
ylab("LDA SCORE (log 10)") +
scale_fill_manual(breaks=c("1", "2"), labels=c("Cluster 1", "Cluster 2"),
                  values=c("1"="#0072B2", "2"="#E69F00"))

print(p)
ggsave("figures/Extended_figure_9h-species_lefse_D0_biontech_JSD_cluster.pdf", p, width=8, height=12)
```

